---
title: "Untitled"
format: html
editor: visual
---

# WORKSHOP 1: - FOUNDATIONS OF DATA SCIENCE - WRANGLING AND PLOTTING
#STEP 1: Clone class repo via R studio
#STEP 2: Remove original connection
#STEP 3: Create new copy to your profile
#Step 4: Verification of files pane and working space

##Housekeeping
## Inventory every active object currently residing in session RAM
objects()
## Purge global environment
rm(list = ls())
# Confirm that global session memory now completely vacant
objects()
#Step 5: breaking the save workspace data

```{r}
##Load the primary data science framework and excel import library
library(tidyverse)
library(readxl)

```

```{r}
#Practice Import A: Loading a standard comma-separated plain text file
benthic_cover <- read_csv(here::here("data/workshop1/reef_cover_log.csv"))
```

```{r}
#Practice Import B: Parsing a tab-separated telemetry instrument array string
acoustic_stream <- read_excel(here::here("data/acoustic_telemtry_sream.txt"))

```

```{r}
#Practice Import C: targeting a specific sheet in a multi-tab excel spreadsheet
fisheries_annual <- read_excel(here::here("data/workshop1/fish_catch_data.xlsx"), sheet = "commercial_2026")
```

```{r}
# Read in mangrove_data
mangrove_data <- read_csv(file = here::here("data/workshop1/mangrove_survey_raw.csv"))

```

```{r}
# Use args within read_csv to skip headers and declare missing flags
mangrove_data <- read_csv(
  here::here("data/workshop1/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA

```

```{r}
# Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover <- as.data.frame(benthic_cover)

```

```{r}
# Print the old-style dataframe structure to view
print(benthic_cover)
# And compare with tibble alternative
print(benthic_cover)

```

```{r}
# Install the data package
# install.packages("palmerpenguins")

# Load the package data
library(palmerpenguins)
data("penguins")

# Examine the structure of the data set - always do this when loading a new data set!
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version

```

```{r}
#Generate an exploratory summary matrix
summary(penguins)
```
```{r}
# Vertically slice specific morphometric variables by explicit name
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)

# Discard logistics tracking attributes while preserving everything else using the minus sign
clean_scientific_fields <- select(penguins, -year)


```
```{r}
# Isolate observations belonging to a single categorical target group
adelie_cohort <- filter(penguins, species == "Adelie")

# Sift out individuals using continuous numerical boundary thresholds
# Preserves only large penguins whose mass exceeds 4500 grams
heavy_penguins <- filter(penguins, body_mass_g > 4500)

# Combine multiple conditional parameters across separate attributes
# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

# Sift records matching multiple targeting flags within an explicit set
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))

```

```{r}
# Sort penguins by ascending body mass (Default setting: Smallest mass first)
lightest_first <- arrange(penguins, body_mass_g)

# Sort penguins in descending sequence using the desc() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))

# Execute nested sorting criteria: Group by species, then sort by descending bill length 
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
```

```{r}
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")

```

```{r}
# Calculate a new morphological ratio in our environment
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )

# View your newly engineered variables appended to the far-right columns
glimpse(penguin_ratios)

```
```{r}
# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)

```

```{r}
# Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)

```
```{r}
#Pipe dataset directly from aggregation to plot with error bars
mass_compare_plot <- penguins |> 
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    se_mass = sd(body_mass_g, na.rm = TRUE)/ sqrt(n()),
    n = n(),
    .groups = "drop"
  )|>
  ggplot(aes(x = species, y = mean_mass, colour = island)) + 
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - se_mass,
                    ymax = mean_mass + se_mass),
                width = 0.2) + 
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()
mass_compare_plot
```
